\chapter{Simple Units and Networks}
\label{ch:simple-units}

\textit{Before we can model neurons, we need to decide what a model is for. This chapter introduces the simplest mathematical neuron---the perceptron---and asks what computations a single unit or a small network can perform. The central ideas are classification, linear separability, and the dot product as a measure of similarity.}

% ============================================================
\section{Why Model?}
\label{sec:why-model}
% ============================================================

A model is a simplified mathematical description of a complex system. The point is not to reproduce every biological detail---that would just be a second copy of the brain, equally hard to understand. The point is to isolate the mechanisms that matter for a particular question and to make predictions that can be tested.

What makes a model ``good'' depends entirely on the question. A model of how ion channels generate an action potential needs biophysical parameters: conductances, gating kinetics, reversal potentials. A model of how the brain classifies visual objects needs none of that---it needs a description of how inputs are weighted and combined. Both are valid. The art is choosing the right level of abstraction.

Computational neuroscience operates across many scales: from the stochastic opening of a single channel, to the electrical dynamics of a single neuron, to the collective activity of thousands of neurons in a network, to the behavior of an animal. At each scale, different state variables and different equations are appropriate. This course begins at the simplest level---a single unit that takes inputs and produces an output---and works upward.

A useful way to organize these scales is by asking what is treated as a state variable and what is treated as hidden detail. At the molecular scale, state variables are channel states and ionic concentrations. At the single-cell scale, they are voltage and gating variables. At the population scale, they are firing rates or spike counts of groups of neurons. At the behavioral scale, they are latent decision variables and motor outputs. The same biological system can therefore admit multiple valid models, each with different observables, equations, and predictive targets.

Computationally, this layered view prevents a common mistake: using a model at the wrong level for the question at hand. We will repeatedly move between levels in later chapters (from channels to spikes to networks), but always with explicit assumptions about what has been coarse-grained away.

% ============================================================
\section{Classification as a Computational Task}
\label{sec:classification}
% ============================================================

Many things the brain does can be framed as classification\index{classification}: given some sensory input, assign it to a category. Is this object a face or not? Is this sound a vowel or a consonant? Should I move left or right?

Abstractly, we have a feature vector $\vec{x} \in S_x$ and a set of labels $S_y$. Classification means finding a function $f: S_x \to S_y$ that maps inputs to correct labels. The challenge is that the space of possible mappings is enormous. If each input is a binary vector of length $n$, there are $2^n$ possible inputs and $|S_y|^{2^n}$ possible functions. Even for modest $n$, this number is far too large for a lookup table.

The brain solves this problem efficiently. There is evidence that higher cortical areas store something like prototypes---internal representations of typical examples for each category. A new input is compared to these prototypes, and the closest match determines the classification. The question is: what is the simplest mathematical device that can do this?

% ============================================================
\section{The Perceptron}
\label{sec:perceptron}
% ============================================================

The perceptron\index{perceptron} is the simplest model of a neuron that can perform classification. It takes $n$ inputs, computes a weighted sum, and passes the result through a nonlinear function:

\begin{keyequation}[The Perceptron]
\begin{equation}
  y = f\!\Bigl(\sum_{i=1}^{n} w_i x_i - \theta\Bigr)
  \label{eq:perceptron}
\end{equation}
\tcblower
$x_i$: input (firing rate of neuron $i$), \quad
$w_i$: synaptic weight, \quad
$\theta$: threshold, \quad
$f$: transfer function.
\end{keyequation}

\noindent The weighted sum $h = \sum_i w_i x_i$ is called the \emph{activation}\index{activation}. The transfer function $f$ determines the output format. Common choices include the step function (McCulloch--Pitts neuron\index{McCulloch--Pitts neuron}), which outputs $1$ if $h > \theta$ and $0$ otherwise, and smooth alternatives like the sigmoid $f(z) = 1/(1 + e^{-z})$.

What does the perceptron actually compute? Writing the activation in vector notation, $h = \vec{w} \cdot \vec{x}$, we see that it is a \emph{dot product}\index{dot product}. The output is large when the input $\vec{x}$ points in roughly the same direction as the weight vector $\vec{w}$, and small or zero when they point in different directions. The perceptron is, at its core, a similarity detector: it measures how closely the current input matches the pattern stored in the weights.

\subsection{Decision boundaries}

The threshold condition $\vec{w} \cdot \vec{x} = \theta$ defines a \emph{hyperplane}\index{hyperplane} in input space. On one side the perceptron outputs $1$; on the other side it outputs $0$. The weight vector $\vec{w}$ is perpendicular to this hyperplane, and the threshold $\theta$ determines how far the hyperplane sits from the origin.

For a two-class problem, the perceptron divides input space into two half-spaces. Any input falling on the $\vec{w}$ side of the hyperplane is classified as ``yes''; anything on the other side is classified as ``no.'' This geometric picture makes the perceptron's power and limitations immediately clear.

% ============================================================
\section{Linear Separability}
\label{sec:linear-sep}
% ============================================================

A classification problem is \emph{linearly separable}\index{linear separability} if there exists a hyperplane that correctly separates all inputs of one class from all inputs of the other. The perceptron can solve a problem if and only if it is linearly separable.

Many problems are linearly separable. If two categories of objects differ systematically along some combination of features, a single hyperplane can separate them. But some problems are not. The canonical example is the XOR function: given two binary inputs $x_1, x_2 \in \{0, 1\}$, output $1$ if exactly one input is $1$, and $0$ otherwise. No single line in the $(x_1, x_2)$ plane can separate the $(0,0)$ and $(1,1)$ points (both class $0$) from the $(0,1)$ and $(1,0)$ points (both class $1$). The perceptron fails here.

This is not a minor technicality. It tells us something fundamental: a single neuron with a linear decision boundary cannot represent arbitrary input--output relationships. To go further, we need either more neurons or a different architecture.

% ============================================================
\section{The Dot Product as Similarity}
\label{sec:dot-product}
% ============================================================

The fact that the perceptron computes a dot product deserves emphasis, because this operation appears throughout neuroscience and machine learning. If both $\vec{w}$ and $\vec{x}$ are normalized to unit length, then $\vec{w} \cdot \vec{x} = \cos\alpha$, where $\alpha$ is the angle between them. The dot product is maximal when the two vectors are aligned and zero when they are orthogonal.

This gives a natural interpretation of what the weights represent: the weight vector $\vec{w}$ is a \emph{template}\index{template matching} or prototype. The perceptron fires when the input is sufficiently similar to the template. This is consistent with the neuroscience idea that certain neurons act as feature detectors---cells in inferotemporal cortex, for instance, respond selectively to specific objects or faces. The weight vector of such a neuron encodes its preferred stimulus.

Winner-take-all competition among multiple perceptrons, each with a different weight vector, implements nearest-prototype classification. The input is assigned to the category whose prototype it most closely resembles. This is one of the simplest models of categorical perception.

% ============================================================
\section{Multilayer Networks}
\label{sec:multilayer}
% ============================================================

The limitations of a single perceptron disappear when we stack multiple layers. A \emph{multilayer feedforward network}\index{multilayer network} passes the input through one or more hidden layers before the output. Each hidden unit computes a weighted sum plus nonlinearity, and the outputs of one layer become the inputs to the next.

Why does depth help? Each hidden unit creates its own hyperplane in input space. The combination of many hyperplanes can carve out arbitrarily complex decision regions. A two-layer network with enough hidden units can separate any set of points---it is not limited to linearly separable problems.

This idea is made precise by the \emph{universal approximation theorem}\index{universal approximation theorem}: a feedforward network with a single hidden layer of sufficient width and a non-polynomial activation function can approximate any continuous function on a compact domain to arbitrary accuracy. The key requirement is that the activation function is nonlinear. A network of purely linear units, no matter how deep, can only compute linear functions---the composition of linear maps is still linear.

The intuition is straightforward. Each hidden unit with a nonlinear activation (say, a ReLU\index{ReLU}) creates a ``kink'' in the function along some hyperplane. By rotating, shifting, and summing many such kinks, the network can sculpt any continuous shape. More hidden units means a finer partition of input space and a closer approximation to the target function.

Of course, the universal approximation theorem says nothing about how to \emph{find} the right weights. That is the problem of learning, which we take up in \cref{ch:learning}.


\begin{chaptersummary}

\begin{center}
\renewcommand{\arraystretch}{1.6}
\begin{tabularx}{\textwidth}{@{}l X X@{}}
  \toprule
  \textbf{Concept} & \textbf{Equation} & \textbf{Key Insight} \\
  \midrule
  Perceptron &
  $\displaystyle y = f\!\Bigl(\sum_i w_i x_i - \theta\Bigr)$ &
  Weighted sum + nonlinearity \\
  %
  Dot product &
  $h = \vec{w} \cdot \vec{x}$ &
  Measures similarity between input and template \\
  %
  Decision boundary &
  $\vec{w} \cdot \vec{x} = \theta$ &
  Hyperplane in input space \\
  %
  Linear separability &
  --- &
  Perceptron solves only linearly separable problems \\
  %
  Universal approximation &
  --- &
  Sufficient width + nonlinearity $\Rightarrow$ arbitrary functions \\
  \bottomrule
\end{tabularx}
\end{center}

\end{chaptersummary}

\begin{conceptquestions}
  \item What makes a model ``good'' in computational neuroscience? Is a model that reproduces every biological detail necessarily better than a simpler one?

  \item A perceptron with weight vector $\vec{w}$ classifies inputs by computing $\vec{w} \cdot \vec{x}$. If two inputs $\vec{x}_1$ and $\vec{x}_2$ produce the same activation $h$, what can you say about their geometric relationship to $\vec{w}$?

  \item The perceptron's decision boundary is the hyperplane $\vec{w} \cdot \vec{x} = \theta$. How does changing $\theta$ (without changing $\vec{w}$) move this boundary? How does changing the magnitude of $\vec{w}$ (without changing its direction or $\theta$) affect classification?

  \item Why can't a single perceptron solve the XOR problem? Give a geometric argument using the decision boundary.

  \item A classification problem has $n$-dimensional binary inputs. How many possible input--output mappings exist? Why does this make a lookup-table approach infeasible even for moderate $n$?

  \item The dot product $\vec{w} \cdot \vec{x}$ can be interpreted as a similarity measure. Under what conditions is this interpretation most natural? What happens to the interpretation when $\vec{w}$ and $\vec{x}$ are not normalized?

  \item Compare the McCulloch--Pitts step function with a smooth sigmoid as the transfer function $f$. What does each choice imply about how the neuron responds to inputs near the decision boundary?

  \item If all activation functions in a multilayer network were linear (i.e., $f(z) = z$), the network could only compute linear functions regardless of its depth. Why? What does this tell you about the role of nonlinearity?

  \item The universal approximation theorem guarantees that a wide enough single-hidden-layer network can approximate any continuous function. Does this mean a single hidden layer is always sufficient in practice? What might deeper networks offer that wider shallow networks do not?

  \item Consider a winner-take-all network of multiple perceptrons, each with a different weight vector. Describe in your own words how this implements nearest-prototype classification. What biological structure might this correspond to?
\end{conceptquestions}
