\chapter{The Membrane Potential}
\label{ch:membrane}

\textit{Every neuron maintains a voltage difference across its membrane---roughly $-70\;\text{mV}$ at rest. This chapter explains why this voltage exists, how we can predict it from first principles, and what happens when we inject current into a cell. The key ideas are diffusion, drift, the Nernst equation, and the membrane as an RC circuit.}

% ============================================================
\section{The Point Neuron Approximation}
\label{sec:point-neuron}
% ============================================================

We begin with a drastic simplification: treat the entire neuron as a single electrical compartment. The membrane is an insulating lipid bilayer separating two conductive solutions. We ignore the spatial extent of dendrites and axons and ask: what determines the voltage $V$ across this membrane? Relaxing this assumption leads to cable theory (\cref{ch:cable}).

The key physical insight is that the lipid bilayer is an insulator sandwiched between two conductors (intra- and extracellular fluid). This is precisely a capacitor. The membrane stores charge, and the voltage across it is determined by the balance of ionic currents flowing through channels and the charge stored on the capacitor.

The fundamental experimental observation is this: if you inject a small current into a neuron, the voltage deflects smoothly from rest. But above a threshold, the response is dramatically different---a sharp, stereotyped spike. This \emph{all-or-nothing response} is the action potential. Understanding why requires understanding the ionic basis of the resting potential.

% ============================================================
\section{Ions and Concentrations}
\label{sec:ions}
% ============================================================

The relevant charge carriers are ions dissolved in the intra- and extracellular fluid. The key players are:

\begin{center}
\begin{tabular}{@{}lccc@{}}
  \toprule
  \textbf{Ion} & \textbf{Valence} $z$ & \textbf{Inside} (mM) & \textbf{Outside} (mM) \\
  \midrule
  $\Na$   & $+1$ & $5$--$15$     & $\sim 145$  \\
  $\K$    & $+1$ & $\sim 140$    & $\sim 5$    \\
  $\Cl$   & $-1$ & $\sim 4$      & $\sim 110$  \\
  $\Ca$   & $+2$ & $\sim 0.1\;\mu\text{M}$ & $2.5$--$5$ \\
  \bottomrule
\end{tabular}
\end{center}

These concentrations are maintained by active transport, particularly the Na$^+$/K$^+$-ATPase pump, which expends energy to move $3\,\Na$ out and $2\,\K$ in per ATP hydrolyzed.

A crucial observation: the bulk solutions on either side of the membrane are very nearly electrically neutral---a property called \emph{space-charge neutrality}. The total positive charge approximately equals the total negative charge in both compartments. The membrane potential arises from a \emph{thin layer} of excess charge at the membrane surface---a tiny fraction of the total ions present.

The Faraday constant $F = N_A q_e \approx 96{,}485\;\text{C\,mol}^{-1}$ converts between moles and coulombs. One mole of monovalent ions carries $F$ coulombs of charge.

% ============================================================
\section{Diffusion: Fick's Law}
\label{sec:diffusion}
% ============================================================

Imagine dropping a blob of dye into water. The dye molecules spread out---not because any force pushes them, but because random thermal motion carries more molecules away from regions of high concentration than toward them. This is diffusion. Each ion undergoes a random walk due to thermal collisions. On average, there is no preferred direction, but in a region with a concentration gradient, more particles randomly step \emph{away} from the dense region than \emph{toward} it, producing a net flux from high to low concentration.

We can derive the diffusion flux from a simple random walk model. Consider a one-dimensional lattice with spacing $\Delta x$. At each time step $\Delta t$, each particle steps left or right with equal probability $1/2$. If the concentration at position $x$ is $C(x)$, the net flux across a boundary is

\begin{keyequation}[Fick's Law]
\begin{equation}
  J_{\text{diff}} = -D \pd{C}{x}
  \label{eq:fick}
\end{equation}
\tcblower
$J_{\text{diff}}$: particle flux $(\text{mol}\,\text{m}^{-2}\text{s}^{-1})$, \quad
$D$: diffusion coefficient $(\text{m}^2/\text{s})$, \quad
$C$: concentration $(\text{mol}/\text{m}^3)$
\end{keyequation}

\noindent where the diffusion coefficient $D = (\Delta x)^2 / (2\Delta t)$. The negative sign captures the essential physics: particles flow \emph{down} the concentration gradient, from high to low concentration.


% ============================================================
\section{Drift in an Electric Field}
\label{sec:drift}
% ============================================================

Ions carry charge, so they respond to electric fields. An electric field $E = -\dd U / \dd x$ exerts a force on each ion, producing a directed drift on top of the random thermal motion. The drift flux for ions of valence $z$ and concentration $C$ is
%
\begin{equation}
  J_{\text{drift}} = -\mu z C \frac{\dd U}{\dd x}
  \label{eq:drift}
\end{equation}
%
where $\mu$ is the \emph{mobility}---how fast ions move per unit force, accounting for friction with the surrounding medium.

Diffusion and drift are not independent phenomena. Both arise from the same thermal fluctuations and friction. The \emph{Einstein relation} connects them:
%
\begin{equation}
  D = \mu \frac{k_B T}{q} = \mu \frac{RT}{zF}
  \label{eq:einstein}
\end{equation}
%
Hotter temperatures mean faster diffusion \emph{and} faster drift for the same applied force. This relation is the key to deriving the Nernst equation.

% ============================================================
\section{The Nernst Equation}
\label{sec:nernst}
% ============================================================

We now ask: at what voltage does the net ionic current vanish for a single ionic species? This is the equilibrium condition where diffusion exactly balances drift.

\subsection{Derivation from zero net flux}

The total flux is the sum of diffusion and drift:
%
\begin{equation}
  J = -D \pd{C}{x} - \mu z C \frac{\dd U}{\dd x}
  \label{eq:total-flux}
\end{equation}

At equilibrium, $J = 0$. Setting the two terms equal and using the Einstein relation $D = \mu RT / (zF)$:

\begin{derivation}[Nernst equation]
Start from $J = 0$:
\begin{align*}
  D \pd{C}{x} &= -\mu z C \frac{\dd U}{\dd x} \\[6pt]
  \frac{RT}{zF} \frac{1}{C}\pd{C}{x} &= -\frac{\dd U}{\dd x}
    && \text{substituting } D = \frac{\mu RT}{zF}
\end{align*}
The left side is $\frac{RT}{zF}\frac{\dd}{\dd x}\ln C$. Integrating across the membrane from inside ($x = 0$) to outside ($x = d$):
\[
  \frac{RT}{zF}\ln\frac{C_{\text{out}}}{C_{\text{in}}} = -(U_{\text{out}} - U_{\text{in}}) = V
\]
where $V = U_{\text{in}} - U_{\text{out}}$ is the membrane potential (intracellular minus extracellular).
\end{derivation}

\begin{keyequation}[The Nernst Equation]
\begin{equation}
  \Erev{X} = \frac{RT}{zF}\ln\frac{\conc{X}{out}}{\conc{X}{in}}
  \label{eq:nernst}
\end{equation}
\tcblower
$R = 8.314\;\text{J\,mol}^{-1}\text{K}^{-1}$: gas constant, \quad
$T$: temperature (K), \quad
$z$: ion valence, \quad
$F = 96{,}485\;\text{C\,mol}^{-1}$: Faraday constant. \\[4pt]
At body temperature $T = 310\;\text{K}$: $\;\;RT/F \approx 26.7\;\text{mV}$.
\end{keyequation}

The quantity $\Erev{\text{ion}}$ is the voltage at which the electrical force on an ion exactly balances its tendency to diffuse down its concentration gradient. If the membrane potential equals the Nernst potential for a given ion, that ion has zero net current---even if channels for that ion are open.

\subsection{The concept of a reversal potential}

The Nernst potential $\Erev{\text{ion}}$ is also called the \emph{reversal potential}\index{reversal potential} because it is the voltage where the ionic current reverses direction. If $V > \Erev{\text{ion}}$, the driving force $(V - \Erev{\text{ion}})$ is positive and a cation channel conducts outward current. If $V < \Erev{\text{ion}}$, the driving force is negative and the channel conducts inward current. At $V = \Erev{\text{ion}}$, no net current flows.

As a concrete example, using $RT/F \approx 26.7\;\text{mV}$ at body temperature: for potassium with $\conc{K}{out} = 5\;\text{mM}$ and $\conc{K}{in} = 140\;\text{mM}$,
%
\begin{align}
  \Erev{K}  &= 26.7\;\text{mV} \times \ln\frac{5}{140}
              \approx -89\;\text{mV},
              \label{eq:erev-k} \\[4pt]
  \Erev{Na} &= 26.7\;\text{mV} \times \ln\frac{145}{10}
              \approx +71\;\text{mV}.
              \label{eq:erev-na}
\end{align}
%
The sign tells us the direction of the equilibrium force: $\K$ wants to pull the cell negative; $\Na$ wants to push it positive. At rest, the membrane is much more permeable to $\K$, so the resting potential ($\approx -70\;\text{mV}$) is close to $\Erev{K}$.

Note that the Nernst equation gives the equilibrium potential for a \emph{single} ionic species, assuming no other ions are permeant. A real membrane is permeable to multiple ions simultaneously, which is why the resting potential is not exactly equal to any single Nernst potential.


% ============================================================
\section{Multiple Ions: The Goldman--Hodgkin--Katz Equation}
\label{sec:ghk}
% ============================================================

A real membrane has channels for several ion types simultaneously. The resting potential depends on all of them, weighted by their permeabilities. The Goldman--Hodgkin--Katz (GHK) voltage equation\index{Goldman--Hodgkin--Katz equation} gives the voltage at which the \emph{total} ionic current vanishes. It assumes a constant electric field across the membrane, non-interacting ions, and steady state (not equilibrium---ions are still flowing, but the net current is zero).

\subsection{Donnan equilibrium versus physiological steady state}

It is important to separate two concepts that are often conflated. In a true thermodynamic equilibrium (Donnan equilibrium\index{Donnan equilibrium}), all ionic fluxes vanish individually and each permeant ion satisfies its own Nernst condition at the same membrane voltage. Real neurons are not in that state. They maintain concentration gradients by active transport (Na/K pump, exchangers), so there are ongoing ionic fluxes and metabolic work.

Resting neurons are better described as a \emph{nonequilibrium steady state}: concentrations are approximately time-invariant, and the \emph{net membrane current} is zero, but individual ionic currents need not vanish. The GHK framework is built precisely for this regime.

\begin{keyequation}[Goldman--Hodgkin--Katz Voltage Equation]
\begin{equation}
  V_{\text{rest}} = \frac{RT}{F}\ln\frac{
    P_{\text{K}}\conc{K}{out} + P_{\text{Na}}\conc{Na}{out} + P_{\text{Cl}}\conc{Cl}{in}
  }{
    P_{\text{K}}\conc{K}{in} + P_{\text{Na}}\conc{Na}{in} + P_{\text{Cl}}\conc{Cl}{out}
  }
  \label{eq:ghk}
\end{equation}
\tcblower
$P_X$: permeability of ion $X$ (cm/s). \quad
$\Cl$ concentrations are flipped (in/out) because $z_{\text{Cl}} = -1$.
\end{keyequation}

The GHK equation is effectively a permeability-weighted average of the individual Nernst potentials. Whichever ion has the highest permeability dominates the resting potential. At rest, $P_{\text{K}} \gg P_{\text{Na}}$, so $V_{\text{rest}} \approx \Erev{K} \approx -70\;\text{mV}$.

\subsection{GHK current equation and nonlinear I--V curves}

The voltage equation above is derived by summing GHK \emph{currents}. For a single ion species $X$ with valence $z_X$ under the constant-field approximation:
%
\begin{keyequation}[Goldman--Hodgkin--Katz Current Equation]
\begin{equation}
  I_X(V) =
  P_X \frac{z_X^2 F^2}{RT}\,V
  \frac{\conc{X}{in} - \conc{X}{out}\exp\!\bigl(-z_XFV/(RT)\bigr)}
       {1 - \exp\!\bigl(-z_XFV/(RT)\bigr)}
  \label{eq:ghk-current}
\end{equation}
\tcblower
$P_X$: permeability, \quad
$z_X$: valence, \quad
$F,R,T$: Faraday constant, gas constant, temperature.
\end{keyequation}
%
\noindent Unlike an Ohmic current, \cref{eq:ghk-current} is nonlinear in $V$. This nonlinearity can produce inward or outward rectification even for a passive membrane with fixed permeabilities, purely from electrodiffusion.

At small voltages ($|z_X F V| \ll RT$), expanding the exponential recovers an approximately linear relation $I_X \approx g_X(V - \Erev{X})$, which explains why Ohmic channel models are often reasonable near rest.

\subsection{An alternative passive nonlinearity: energy barriers}

A second route to non-Ohmic behavior is an internal energy barrier in the channel pore (Eyring rate theory\index{Eyring rate theory}). In the simplest one-barrier model, inward and outward hopping rates depend asymmetrically on membrane voltage:
%
\begin{equation}
  I \propto
  \conc{X}{in}\,\exp\!\Bigl(-\delta \frac{zFV}{RT}\Bigr)
  -
  \conc{X}{out}\,\exp\!\Bigl((1-\delta)\frac{zFV}{RT}\Bigr),
  \label{eq:eyring}
\end{equation}
%
where $\delta \in [0,1]$ is the fractional barrier position within the membrane field. When $\delta=1/2$ the barrier is symmetric; $\delta \neq 1/2$ yields asymmetric rectification.

The key conceptual point is that passive nonlinear I--V curves need not imply voltage-gated channel dynamics. They can arise from electrodiffusion (GHK) or from asymmetric barrier crossing (Eyring) even before active gating is introduced.

Computationally, this matters because the baseline membrane transfer function may already be nonlinear before we add voltage-dependent channels, altering gain and subthreshold filtering.

\subsection{Why extracellular potassium depolarizes neurons}

If we increase $\conc{K}{out}$ (a common clinical scenario in hyperkalemia), the Nernst potential $\Erev{K}$ becomes less negative because the log ratio $\conc{K}{out}/\conc{K}{in}$ increases. Since the resting potential is dominated by $\K$ permeability, the resting potential moves toward zero---the cell depolarizes. For example, raising $\conc{K}{out}$ from the normal $5\;\text{mM}$ to $15\;\text{mM}$ shifts $\Erev{K}$ from $-89\;\text{mV}$ to roughly $-60\;\text{mV}$, a significant depolarization that can make neurons hyperexcitable.

% ============================================================
\section{The Membrane as an RC Circuit}
\label{sec:rc-circuit}
% ============================================================

We now model the membrane as an electrical circuit. This is the foundation for every dynamical neuron model in this course.

\subsection{Circuit components}

The membrane has three electrical properties: a \textbf{capacitance} $\Cm$ (the lipid bilayer stores charge; typical value $\sim 1\;\mu\text{F}/\text{cm}^2$), \textbf{conductances} $g_i$ (ion channels act as resistors, each with its own reversal potential), and an \textbf{external current} $I_{\text{ext}}$ (injected by an electrode or, physiologically, by synapses).

\begin{center}
\input{tikz/circuits}
\begin{tikzpicture}[scale=0.9, every node/.style={font=\small}]
  % Capacitor branch (leftmost)
  \draw (0,0) node[above] {inside}
    to[capacitor, l=$\Cm$] (0,-3)
    node[below] {outside};

  % Leak branch
  \draw (2,0) -- (2,0)
    to[american resistor, l=$g_L$] (2,-1.5)
    to[battery1, l_=$\Erev{L}$] (2,-3);

  % K branch
  \draw (4,0) -- (4,0)
    to[variable american resistor, l=$g_K$] (4,-1.5)
    to[battery1, l_=$\Erev{K}$] (4,-3);

  % Na branch
  \draw (6,0) -- (6,0)
    to[variable american resistor, l=$g_{Na}$] (6,-1.5)
    to[battery1, l_=$\Erev{Na}$] (6,-3);

  % External current source
  \draw (8,0) -- (8,0)
    to[american current source, l=$I_{\text{ext}}$] (8,-3);

  % Top and bottom rails
  \draw (-0.5,0) -- (8.5,0);
  \draw (-0.5,-3) -- (8.5,-3);
\end{tikzpicture}
\end{center}

\subsection{Kirchhoff's current law}

By Kirchhoff's law, the total current into the cell must equal the sum of all currents leaving through channels plus the current charging the capacitor. We adopt the sign convention that positive current flows \emph{into} the cell, and ionic currents are positive when flowing outward.

\begin{keyequation}[The Membrane Equation]
\begin{equation}
  \Cm\frac{\dd V}{\dd t} = -\sum_i g_i\bigl(V - \Erev{i}\bigr) + I_{\text{ext}}(t)
  \label{eq:membrane}
\end{equation}
\tcblower
$\Cm$: membrane capacitance, \quad
$g_i$: conductance of channel type $i$, \quad
$\Erev{i}$: reversal potential, \quad
$I_{\text{ext}}$: injected current.
\end{keyequation}

Each term has a clear physical meaning. The left side $\Cm\,\dd V/\dd t$ is the rate of charge accumulation on the capacitor. Each $-g_i(V - \Erev{i})$ term is the current through channel type $i$; the factor $(V - \Erev{i})$ is the \emph{driving force}, which vanishes at the reversal potential, reverses sign across it, and grows linearly away from it. The term $I_{\text{ext}}$ is whatever current we inject.

\subsection{Single-conductance case: the passive membrane}

The simplest case has a single leak conductance $\gleak$ with reversal potential $\Erev{L} \approx \Vrest$:
%
\begin{equation}
  \Cm\frac{\dd V}{\dd t} = -\gleak\bigl(V - \Erev{L}\bigr) + I_{\text{ext}}(t)
  \label{eq:passive-membrane}
\end{equation}
%
This defines the membrane time constant and input resistance:
%
\begin{equation}
  \taum = \frac{\Cm}{\gleak}\,, \qquad R_m = \frac{1}{\gleak}
  \label{eq:tau-Rm}
\end{equation}
%
For typical values $\Cm = 200\;\text{pF}$ and $\gleak = 10\;\text{nS}$, we get $\taum = 20\;\text{ms}$ and $R_m = 100\;\text{M}\Omega$. The time constant sets the temporal scale of subthreshold voltage dynamics.

\subsection{Response to a constant current step}

Suppose $V(0) = \Erev{L}$ and we inject a constant current $I_0$ starting at $t = 0$. Dividing \cref{eq:passive-membrane} by $\gleak$:
%
\begin{equation}
  \taum \frac{\dd V}{\dd t} = -\bigl(V - \Erev{L}\bigr) + R_m I_0
  \label{eq:passive-rewritten}
\end{equation}
%
This is a first-order linear ODE. The solution is an exponential approach to a new steady state:
%
\begin{equation}
  V(t) = \Erev{L} + R_m I_0 \Bigl(1 - e^{-t/\taum}\Bigr)
  \label{eq:step-response}
\end{equation}
%
The voltage rises exponentially from $\Erev{L}$ toward the steady-state value $V_\infty = \Erev{L} + R_m I_0$ with time constant $\taum$. This exponential charging is precisely the subthreshold dynamics of the leaky integrate-and-fire neuron (\cref{ch:lif}). Adding a spike threshold and reset to \cref{eq:passive-membrane} gives the LIF model---the simplest spiking neuron. The membrane time constant $\taum$ sets how long the neuron ``remembers'' its input.

Note that adding more open channels \emph{decreases} the time constant: the membrane responds faster but the voltage excursions become smaller, since $R_{\text{eff}} = 1/g_{\text{tot}}$ decreases. Inhibition with $\Erev{\text{inh}} \approx \Vrest$ does not hyperpolarize the cell, but it ``shunts'' excitatory drive by dividing the input resistance. This is called \emph{shunting inhibition}\index{shunting inhibition}.


\begin{chaptersummary}

\begin{center}
\renewcommand{\arraystretch}{1.6}
\begin{tabularx}{\textwidth}{@{}l X X@{}}
  \toprule
  \textbf{Concept} & \textbf{Equation} & \textbf{Key Insight} \\
  \midrule
  Fick's law &
  $\displaystyle J_{\text{diff}} = -D\,\pd{C}{x}$ &
  Particles flow down concentration gradient \\
  %
  Einstein relation &
  $\displaystyle D = \frac{\mu RT}{zF}$ &
  Links diffusion to drift \\
  %
  Nernst equation &
  $\displaystyle \Erev{X} = \frac{RT}{zF}\ln\frac{\conc{X}{out}}{\conc{X}{in}}$ &
  Zero-current voltage for one ion \\
  %
  GHK equation &
  $\displaystyle V = \frac{RT}{F}\ln\frac{P_K\conc{K}{out}+\cdots}{P_K\conc{K}{in}+\cdots}$ &
  Permeability-weighted multi-ion potential \\
  %
  Membrane equation &
  $\displaystyle \Cm\dot{V} = -\sum g_i(V-\Erev{i}) + I_{\text{ext}}$ &
  RC circuit with ionic batteries \\
  %
  Time constant &
  $\displaystyle \taum = \Cm / g_{\text{tot}}$ &
  Timescale of voltage dynamics \\
  %
  Resting potential &
  $\displaystyle V_{\text{rest}} = \frac{\sum g_i \Erev{i}}{\sum g_i}$ &
  Conductance-weighted average \\
  \bottomrule
\end{tabularx}
\end{center}

\end{chaptersummary}

\begin{conceptquestions}
  \item The lipid bilayer is an insulator between two conductors. Why does this make it a capacitor? What determines the capacitance?

  \item Diffusion arises from random thermal motion, yet it produces a directed net flux. Explain this apparent paradox using the concept of a concentration gradient.

  \item Why does Fick's law have a negative sign? What would happen physically if the sign were positive?

  \item The Einstein relation $D = \mu k_B T / q$ links diffusion and drift. What does it mean physically that both depend on temperature?

  \item Why does the Nernst equation have a logarithm? What physical balance does it represent? Walk through the derivation from setting the total flux to zero.

  \item The Nernst potential for $\K$ is approximately $-89\;\text{mV}$ and for $\Na$ is approximately $+71\;\text{mV}$. Explain the sign of each in terms of the concentration gradients.

  \item The resting potential of a typical neuron ($\approx -70\;\text{mV}$) is close to $\Erev{K}$ but not exactly equal. Why not?

  \item A patient has dangerously elevated blood potassium ($\conc{K}{out} = 8\;\text{mM}$ instead of $5\;\text{mM}$). Without computing exact numbers, explain qualitatively what happens to the resting potential and why this is dangerous.

  \item Why is the Nernst equation an \emph{equilibrium} result while the GHK equation describes a \emph{steady state}? What is the difference?

  \item In the GHK equation, the $\Cl$ concentrations appear with inside and outside reversed compared to the cations. Why? What would happen if you forgot to flip them?

  \item In the membrane equation, what is the physical meaning of each term: $\Cm\,\dd V/\dd t$, $-g_i(V - \Erev{i})$, and $I_{\text{ext}}$?

  \item The driving force $(V - \Erev{i})$ can be positive, negative, or zero. What does each case correspond to physically? What happens to the current through channel type $i$ when $V = \Erev{i}$?

  \item In the membrane equation, what happens to the voltage dynamics if you double all conductances simultaneously? Does the resting potential change? Does the time constant change?

  \item The step response of a passive membrane is $V(t) = \Erev{L} + R_m I_0(1 - e^{-t/\taum})$. Explain qualitatively why the voltage approaches a new steady state exponentially rather than linearly. What determines the final value?

  \item Explain shunting inhibition in terms of the membrane equation. Why does adding an inhibitory conductance with $\Erev{\text{inh}} \approx \Vrest$ reduce the response to excitation without hyperpolarizing the cell?
\end{conceptquestions}
